\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{A Radial Time Projection Chamber for $\alpha$ detection in CLAS at JLab}

### 2017-05-18 eg6-letters/NIM/EG6_RTPC_NIM.tex
Added changes from Andrea

Measurement of the Deeply Virtual Compton Scattering (DVCS) process
($eH \rightarrow e' H' \gamma$, where $H$ is a hadron) necessitates an upgrade
of this setup.  With a 6 GeV electron beam, the majority of DVCS photons are produced
at very forward angles, where the acceptance of the EC is poor. To extend the detection range,
an inner calorimeter (IC) was built for the E01-113 experiment in 2005.
The IC is constructed from 424 lead-tungstate (PbWO$_{4}$) crystals, covering polar 
angles between 5$^{\circ}$ and 15$^{\circ}$. Each crystal is 16 cm long, corresponding
to 17 radiation lengths. The achieved energy resolution is around 3 to 4$\%$ for photon
energies between 2 GeV and 5 GeV, with an angular resolution between 3 and 5 mrad 
\cite{Hyon-suk}. To protect this detector from the large flux of low energy M{\o}ller 
electrons, a 5~T solenoid was placed around the target to shield the IC. 
In addition, the recoil nuclei of the coherent DVCS on helium was detected 
using a RTPC developed to track low energy nuclear fragments. The solenoid field
serving as the analyzing magnet for the particle tracking in the RTPC. This 
setup was used during a three months experimental run
in 2009 with a longitudinally polarized electron beam of 130 nA 
and energy of 6.064 GeV incident on a gaseous $^{4}$He target.
{\bf \color{red} Add luminosity.}

In this paper we report the design, calibration, and performance of the radial 
TPC, organized as follows. In section \ref{sec_design}, we detail the design 
and internal structure of the detector. Section \ref{sec_readout} is focused 
on the properties of the read-out system. Calibration 
strategies are discussed in section \ref{sec_calib}. Finally, the performances
of the RTPC are described in section \ref{sec_perf}.

@@

The RTPC electron collection system has 3200 readout pads. These elements are
located at the end of the amplification region, 69 mm from the central axis.
Figure \ref{fig:PADs} shows a schematic drawing of the size and 
configuration of the pads. Each readout pad is 5 mm in length and 4.45 mm in 
width.  The shift between the rows allows to reduce aliasing. Each half of the 
RTPC has 40 rows and 40 columns of pads. The shaded region in figure~\ref{fig:PADs}
shows how pads are grouped to 16 channels pre-amplifiers. A Time to Digital
Converter (TDC) with 114~ns samples is used to measure the time taken by
the electrons to drift from the ionization point to the readout board.  Signal
amplitude is is measured with an Analog-to-Digital-Converter (ADC) without
specific normalization.

{\bf \color{red} 
Andrea:

- Mention ALTRO chip and reference.

- Sparcification method (3 consecutive ADC samples above a threshold of xxx ?, and three
samples before and after)

- Mention trigger time T=15.

- Preamplifiers were from Bonus ?

- Basics of JLab fADC (is there a reference?)

- Mention max rate we ran

- Other relevant information ?
}

@@

>>>>

Measurement of the Deeply Virtual Compton Scattering (DVCS) process
($eH \rightarrow e' H' \gamma$, where $H$ is a hadron) necessitates an upgrade
of this setup.  With a 6 GeV electron beam, the majority of DVCS photons are produced
at very forward angles, where the acceptance of the EC is poor. To extend the detection range,
an inner calorimeter (IC) was built for the E01-113 experiment in 2005.
The IC is constructed from 424 lead-tungstate (PbWO$_{4}$) crystals, covering polar 
angles between 5$^{\circ}$ and 15$^{\circ}$. Each crystal is 16 cm long, corresponding
to 17 radiation lengths. The achieved energy resolution is around 3 to 4$\%$ for photon
energies between 2 GeV and 5 GeV, with an angular resolution between 3 and 5 mrad 
\cite{Hyon-suk}. To protect this detector from the large flux of low energy M{\o}ller 
electrons, a 5~T solenoid was placed around the target to shield the IC. 
In addition, the recoil nuclei of the coherent DVCS on helium was detected 
using a RTPC developed to track low energy nuclear fragments. The solenoid field
serving as the analyzing magnet for the particle tracking in the RTPC. This 
setup was used during a three months experimental run
in 2009 with a longitudinally polarized electron beam of 130 nA 
and energy of 6.064 GeV incident on a gaseous $^{4}$He target.

In this paper we report the design, calibration, and performance of the radial 
TPC, organized as follows. In section \ref{sec_design}, we detail the design 
and internal structure of the detector. Section \ref{sec_readout} describes 
the properties of the read-out system. Calibration 
strategies are discussed in section \ref{sec_calib}. Finally, the performances
of the RTPC are described in section \ref{sec_perf}.

@@

The RTPC electron collection system has 3200 readout pads. These elements are
located at the end of the amplification region, 69 mm from the central axis.
Figure \ref{fig:PADs} shows a schematic drawing of the size and 
configuration of the pads. Each readout pad is 5 mm in length and 4.45 mm in 
width.  The shift between the rows allows to reduce aliasing. Each half of the 
RTPC has 40 rows and 40 columns of pads. The shaded region in figure~\ref{fig:PADs} 
shows how pads are grouped to 16 channels pre-amplifiers. The pre-amplifier boards, 
already employed in the BoNuS RTPC \cite{BONUS-NIM}, serve the dual purpose of 
inverting the RTPC signals polarity -- from negative to positive -- to match the 
requirements of the subsequent readout system, and driving the 6m long ribbon 
cable that connects to it.

The readout system is based on the Front End Electronic (FEE) boards originally 
developed for the ALICE TPC readout system at Cern~\cite{ALICE-FEE}. Each 
readout channel is made of three main components: a charge sensitive, low 
impedance shaping amplifier, a 10 bit, 25-MHz digitizer, and a digital circuit 
that implements online-processing algorithms (pedestal subtraction, zero 
suppression, ad tail cancellation). The amplifier is implemented in a 
fully-analogue ASIC, PASA~\cite{ALICE-PASA}, while the subsequent signal 
digitization and elaboration part is implemented in a digital ASIC, 
ALTRO~\cite{ALICE-ALTRO}. Both ASICs host 16 channels. The full readout 
chain is implemented on a single Front End Card (FEC), hosting 128 channels. 
FECs are hosted in custom crates, mounted close to the detector. A Readout 
Control Unit (ROC) board is used to distribute the trigger signal to FECs 
and for data readout. Each RCU can handle up to 25 FECs. Communication 
between ROC and FECs is performed trough a custom backplane, implementing 
a low-voltage signal bus. The RCU communicates with the CLAS DAQ system 
trough a 200 MB/s optical link, connected to a data acquisition PC hosting 
a ReadOut Receiver Card (RORC). This PC hosts the readout controller 
application (ROC) used to interface with the CLAS DAQ system. An Ethernet 
link is also present, for slow-controls and monitoring.

Triggering and data-readout scheme is as follows. When a Level-1 (L1) trigger 
is received by a FEC-board, a programmable number of samples $N_0$ is digitized 
and stored in a data memory, and hence processed by the digital circuit. Upon 
arrival of Level-2 trigger, the latest processed samples are saved in a 
multi-event memory, invididual to each-channel, ready for readout by the RCU. 
In our setup, the RCU receives the trigger signal from the CLAS Trigger 
Supervisor system, and distributes both L1 and L2 signals to all the FECs in 
the crate, with a definite L1-L2 trigger latency. Such latency is programmable, 
and has to be longh enough to accomodate the digitization of the $N_0$ samples 
per channel. In the current implementation, $N_0=100$ samples/channel are 
acquired for each trigger, operating ALTRO at the reduced sampling frequency 
of $10 MHz$. The L1-L2 latency is thus fixed to 15 $\mu$s.

The trigger signal also initiates RCU readout operation from FEC boards. All 
the measured samples from active channels are reported, together with a channel 
identifier and a timestamp, to the ROC application, that in turns sends them on 
the main event builder. During data reconstruction, the acquired samples are 
processed to obtain, for each readout pad, the accumulated charge (ADC) and 
the pulse time (T). Since pulse time was obtained as the time-stamp of the 
first sample above threshold, referred to to the trigger time, the resolution 
is equivalent to the ALTRO sampling time, 100~ns.

\begin{figure}[tb]
   \centering
   \includegraphics[width=.5\textwidth]{fig/ReadoutScheme.png}
   \caption[]{Schematic representation of the RTPC readout system, showing} 
   \label{fig:ReadoutScheme}
\end{figure}

In order to reduce the data size, ALTRO is operated in zero-suppression mode: 
only samples above a programmable threshold are stored in the ALTRO on-board 
memory and then written to tape. A glitch-filter permits to reject spurious 
pulses due to noise, by requiring the presence of a minimum number of samples 
over threshold $N_{MINSEQ}$ to validate a sample. To properly reconstruct 
the signal shape, $N_{PRE}$ samples before threshold-crossing and $N_{POST}$ 
samles after the signal returns below threshold are saved. In the present 
configuration, $N_{MINSEQ}=N_{PRE}=N_{POST}=3$, while threshold is set 
just above the noise level.

In order to read all the detector readout pads, four FEC crates are used, 
each equipped with 6 boards, plus a ROC. A schematic of the readout system, 
for a single crate, is reported in figure~\ref{fig:ReadoutScheme}. This 
configuration permits to reduce the dead time associated to FEC readout 
operations, scaling linearly with the number of boards in the crate. During 
the 2009 run, the system was succesfully operated with a DAQ rate of 3.1 kHz, 
with a live time of $70 \%$.

@@

 \bibitem{ALICE-FEE}
 L. Musa et al., The ALICE TPC front end electronics, 2003 IEEE Nuclear 
 Science Symposium, 2003, pp. 3647-3651 Vol.5

 \bibitem{ALICE-PASA}
   H.~K.~Soltveit {\it et al.},
   ``The Preamplifier shaper for the ALICE TPC-Detector,''
   Nucl.\ Instrum.\ Meth.\ A {\bf 676} (2012) 106

 \bibitem{ALICE-ALTRO},
 R. E. Bosch, {\it et al.}, ``The ALTRO chip: a 16-channel A/D converter and 
 digital processor for gas detectors'', IEEE Transactions on Nuclear Science, 
 vol. 50, no. 6, pp. 2460-2469, Dec. 2003.
